---
title: "Euclidean Cloud"
author: "Cory Brunson"
date: "November 6, 2015"
output: pdf_document
---

This vignette (loosely or closely?) follows the treatment in Chapter 3 of [Le Roux and Rouanet's *Geometric Data Analysis*](http://www.springer.com/us/book/9781402022357), with drastically simplified exposition but employing the same example data.

```{r, echo=FALSE, results=FALSE}
devel = TRUE
if (devel) {
    library(devtools)
    load_all()
} else {
    library(cloud)
}
```

The central object of GDA is a point cloud in a Euclidean space, which encodes a set of statistical observations arranged in a table, or two-dimensional array (for instance, a contingency table in Correspondence Analysis or an Individuals--Variables table in Principal Components Analysis). By convention, the columns of the table correspond to the coordinate axes of the Euclidean space while the rows correspond to the points.

## 3.1 Basic Statistics

Rigorously, a *point cloud* in a Euclidean space $\mathcal{U}$ consists of a set $J$ of labels and a mapping $M^J:J\to\mathcal{U}$ that takes each label $j\in J$ to its corresponding point $M^j\in\mathcal{U}$.

As a running example, we load the *Target example* dataset, used throughout the chapter, whose coordinates are provided in Exercise 3.4:

```{r}
data(Target)
print(Target)
class(Target)
```

```{r, echo=FALSE, results=FALSE}
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     Target, pch = 16,
     xlab = "", ylab = "")
text(Target, labels = rownames(Target), pos = 1)
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
legend(x = -19, y = 19, box.lty = 0, legend = "Target example")
```

The labels consist of $j_1$ through $j_{10}$. Because these data occupy two dimensions, i.e. the labels are mapped into $\mathcal{U}=\mathbb{R}^2$, they constitute a *plane cloud*.

### 3.1.1 Mean Point

The *mean point* of the cloud $M^J$ in $\mathcal{U}$ is meant to be a proxy point location for the entire cloud. Given an arbitrary point $P\in\mathcal{U}$, the mean point is defined as the point arrived to from $P$ via the sum of the vectors $\overrightarrow{PM^j}$ from $P$ to each point $M^j$---that is, as the (unique!) point $G\in\mathcal{U}$ for which $$\overrightarrow{PG}=\sum_{j\in J}\overrightarrow{PM^j}$$ doesn't depend on the choice of point $P\in\mathcal{U}$. The mean point satisfies the *Barycentric property* that $$\sum_{j\in J}\overrightarrow{GM^j}=\overrightarrow{0},$$ which can be checked by substituting $P=G$ in the definition.[^1]

```{r}
Target_barycenter <- barycenter(Target)
print(Target_barycenter)
```

```{r, echo=FALSE, results=FALSE}
# Vectors from barycenter to each point
Target_vectors <- Target - do.call(rbind,
                                   replicate(nrow(Target),
                                             Target_barycenter,
                                             simplify = FALSE))
# Plot barycenter and vectors therefrom to points in cloud
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     x = c(), y = c(),
     xlab = "", ylab = "")
text(Target_barycenter, labels = "G", adj = c(1.5, 1.25))
for (i in 1:nrow(Target)) {
    text(Target[i, , drop = FALSE], labels = rownames(Target)[i],
         adj = .5 - (Target_vectors[i, ] / sqrt(sum(Target_vectors[i, ] ^ 2))))
}
arrows(x0 = Target_barycenter[, 1], y0 = Target_barycenter[, 2],
       x1 = Target[, 1], y1 = Target[, 2],
       length = .2, angle = 15)
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
legend(x = -19, y = 19, box.lty = 0,
       legend = "Target: barycentric characterization")
```

[^1]: These and other formulas are given in more general terms in *GDA*, in which the points $M^j$ are weighted by masses $\omega_j$. I may be forced to adopt this generality, but i haven't yet.