---
title: "Euclidean Cloud"
author: "Cory Brunson, from Brigitte Le Roux and Henry Rouanet"
date: "November 6, 2015"
output: pdf_document
---

This vignette (loosely or closely?) follows the treatment in Chapter 3 of [Le Roux and Rouanet's *Geometric Data Analysis*](http://www.springer.com/us/book/9781402022357), with drastically simplified exposition but employing the same example data.

```{r, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
library(devtools)
devel = TRUE
if (devel) {
    load_all("~/Documents/edu/R/cloud")
} else {
    install_github("corybrunson/cloud")
    library(cloud)
}
library(knitr)
opts_chunk$set(echo=FALSE, results=FALSE, message=FALSE, warning=FALSE)
```

The central object of GDA is a point cloud in a Euclidean space, which encodes a set of statistical observations arranged in a table, or two-dimensional array (for instance, a contingency table in Correspondence Analysis or an Individuals--Variables table in Principal Components Analysis). By convention, the columns of the table correspond to the coordinate axes of the Euclidean space while the rows correspond to the points.

## 3.1 Basic Statistics

Rigorously, a *point cloud* in a Euclidean space $\mathcal{U}$ consists of a set $J$ of labels and a mapping $M^J:J\to\mathcal{U}$ that takes each label $j\in J$ to its corresponding point $M^j\in\mathcal{U}$. The points are assigned positive weights $\omega_j$, which may be assumed unitary when not specified. The most common weights will be absolute frequencies $n_j$, $n=\sum_{j\in J}n_j$, and their associated relative frequencies $f_j=\frac{n_j}{n}$.

As a running example, we load the *Target example* dataset, used throughout the chapter, whose coordinates are provided in Exercise 3.4:

```{r, echo=TRUE}
data(Target)
print(Target)
print(class(Target))
T_freq <- rep(1 / nrow(Target), nrow(Target))
```

```{r}
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     Target, pch = 16,
     xlab = "", ylab = "")
text(Target, labels = rownames(Target), pos = 1)
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
legend(x = -19, y = 19, box.lty = 0, legend = "Target example")
```

The labels consist of $j_1$ through $j_{10}$. Because these data occupy two dimensions, i.e. the labels are mapped into $\mathcal{U}=\mathbb{R}^2$, they constitute a *plane cloud*.

### 3.1.1 Mean Point

The *mean point* of the cloud $M^J$ in $\mathcal{U}$ is meant to be a proxy point location for the entire cloud. Given an arbitrary point $P\in\mathcal{U}$, the mean point is defined as the point arrived to from $P$ via the sum of the weighted vectors $f_j\overrightarrow{PM^j}$ from $P$ to each point $M^j$---that is, as the (unique!) point $G\in\mathcal{U}$ for which $$\overrightarrow{PG}=\sum_{j\in J}f_j\overrightarrow{PM^j}$$ doesn't depend on the choice of point $P\in\mathcal{U}$. The mean point satisfies the *Barycentric property* that $$\sum_{j\in J}f_j\overrightarrow{GM^j}=\overrightarrow{0},$$ which can be checked by substituting $P=G$ in the definition.[^1]

```{r, echo=TRUE}
G <- barycenter(Target)
print(G)
```

```{r}
# Vectors from barycenter to each point
T_vectors <- Target - do.call(rbind,
                                   replicate(nrow(Target),
                                             G,
                                             simplify = FALSE))
# Plot barycenter and vectors therefrom to points in cloud
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     x = c(), y = c(),
     xlab = "", ylab = "")
text(G, labels = "G", adj = c(1.5, 1.25))
for (i in 1:nrow(Target)) {
    text(Target[i, , drop = FALSE], labels = rownames(Target)[i],
         adj = .5 - (T_vectors[i, ] / sqrt(sum(T_vectors[i, ] ^ 2))))
}
arrows(x0 = G[, 1], y0 = G[, 2],
       x1 = Target[, 1], y1 = Target[, 2],
       length = .15, angle = 15)
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
legend(x = -19, y = 19, box.lty = 0,
       legend = "Target: barycentric characterization")
```

[^1]: These and other formulas are given in more general terms in *GDA*, in which the points $M^j$ are weighted by masses $\omega_j$. I may be forced to adopt this generality, but i haven't yet.

### 3.1.2 Inertia, Sum of Squares, Variance and Contributions

The *inertia* of a cloud $M^J$ with respect to a point $P$ is defined, as in other settings, as the sum-squared distance of each point $M^j$ from $P$: $${In}^PM^J=\sum_{j\in J}{In}_j^P=\sum_{j\in J}\omega_j(PM^j)^2.$$ The inertia for $\omega_j=n_j$ is the *sum of squares* and that for $\omega_j=f_j$ is the *mean of squares*. The *variance* of the cloud is its mean of squares from its barycenter, and can be interpreted as the sum of the *absolute contributions* of the individual points in the cloud: $${Var}M^J=\sum_{j\in J}{Cta}_j=\sum_{j\in J}f_j(GM^j)^2.$$ The *relative contributions of the points are, then, the fractions ${Ctr}_j=\frac{{Cta}_j}{{Var}M^J}$.

```{r, echo=TRUE}
P <- c(0, 0)
P_in <- inertia(point = P, cloud = Target)
G_in <- inertia(point = G, cloud = Target)
T_var <- inertia(G, cloud = Target,
                           weights = T_freq)
print(c(center_inertia = P_in,
        barycenter_inertia = G_in,
        variance = T_var))
```

```{r, echo=TRUE}
Ctas <- T_freq * apply(Target, 1, function(m) {
    sum((m - G) ^ 2)
})
Ctrs <- Ctas / T_var
print(cbind(Ctas, Ctrs))
```

The **first Huyghen's theorem** states that the inertia of a cloud $M^J$ with respect to a point $P$ can be decomposed into the sum of the variance of the cloud (its inertia with respect to its barycenter $G$) and the squared distance between $G$ and $P$: $$\sum_{j\in J}f_j(PM^j)^2=(PG)^2+\sum_{j\in J}f_j(GM^j)^2.$$ (See the illustration below.) A consequence is the *metric characterization of the barycenter*, which has that the barycenter minimizes the mean of squares of the cloud. That is, across $P\in\mathcal{U}$, the quantity $\sum_{j\in J}f_j(PM^j)^2$ is minimized when $P=G$.

```{r, echo=TRUE}
huyghen.test(point = P, cloud = Target, weights = T_freq)
huyghen.test(point = G, cloud = Target, weights = T_freq)
```

```{r}
# Three plots!
par(mfrow = c(1, 3))
# Plot points relative to center
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     Target, pch = 16,
     xlab = "", ylab = "")
segments(x0 = 0, y0 = 0, x1 = Target[, 1], y1 = Target[, 2])
text(x = P[1], y = P[2], labels = "P",
     adj = c(sqrt(2), sqrt(2)))
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
# Plot center relative to barycenter
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     x = c(), y = c(),
     xlab = "", ylab = "")
points(rbind(P, G),
       pch = c(13, 21), bg = c("white", "grey"), cex = 2)
segments(x0 = P[1], y0 = P[2],
         x1 = G[1], y1 = G[2])
text(x = rbind(P, G),
     labels = c("P", "G"), adj = c(1, -1))
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
# Plot points relative to barycenter
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     Target, pch = 16,
     xlab = "", ylab = "")
segments(x0 = G[1], y0 = G[2],
         x1 = Target[, 1], y1 = Target[, 2])
text(x = G[1], y = G[2], labels = "G",
     adj = c(sqrt(3), 1))
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
# Reset
par(mfrow = c(1, 1))
```

## 3.2 Projected Clouds

The *(orthogonal) projection* of a point cloud $M^J$ in a space $\mathcal{U}$ onto a subspace $\mathcal{H}$ is the cloud $H^J=\{H^j\}_{j\in J}$ consisting of its points' projections, and called the *projected cloud*. An important property of the projected cloud is that its barycenter $G'$ is the projection of the barycenter $G$ of $M^J$.

### 3.2.1 Variance in a Direction

The variances of the projected clouds of $M^J$ onto two equidimensional parallel subspaces $\mathcal{H}$ and $\mathcal{H}'$ are equal, and are called the *variance in the direction* of $\mathcal{H}$ (or of $\mathcal{H}'$). A special case is the variance in the direction of a line, which is equal to $$\sum_{j\in J}f_j\frac{\langle\overrightarrow{GM}^j\mid\overrightarrow{\alpha}\rangle^2}{\|\overrightarrow{\alpha}\|^2}$$ and is called the *variance of axis*.

```{r, echo=TRUE}
axis <- rbind(c(0, 0), c(2, 1))
axis_var <- direction.variance(cloud = Target, weights = T_freq,
                               subspace = axis)
print(axis_var)
print(direction.variance(cloud = Target, weights = T_freq,
                         subspace = affine.decomposition(axis)$linear.subspace,
                         type = "linear"))
```

```{r}
# Three plots!
par(mfrow = c(1, 2))
# Plot cloud, axis, and residual segments
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     Target, pch = 16,
     xlab = "", ylab = "")
points(x = P[1], y = P[2], pch = 13, cex = 2)
m <- diff(axis[, 2]) / diff(axis[, 1])
abline(a = axis[1, 2] - m * axis[1, 1], b = m, lty = 2)
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
# Plot center relative to barycenter
plot(xlim = c(-19, 19), ylim = c(-19, 19), axes = FALSE, asp = 1,
     x = c(), y = c(),
     xlab = "", ylab = "")
points(rbind(P, G),
       pch = c(13, 21), bg = c("white", "grey"), cex = 2)
segments(x0 = P[1], y0 = P[2],
         x1 = G[1], y1 = G[2])
text(x = rbind(P, G),
     labels = c("P", "G"), adj = c(1, -1))
rect(xleft = -19, xright = 19, ybottom = -19, ytop = 19, lty = 3)
# Plot cloud, new axis, and new residual segments

# Reset
par(mfrow = c(1, 1))
```

### 3.2.2 Residual Square Mean

The first Huyghen's theorem can be extended from "point" variance to directional variance by way of projection: Given a cloud $M^J$ and a subspace $\mathcal{H}$, the *residual square mean* $$\sum_{j\in J}f_j(M^jH^j)^2$$ of $M^J$ with respect to $\mathcal{H}$ is the weighted mean of the squares of the "residuals" $M^jH^j$---the distances from the points of $M^J$ to their projections in $H^J$. The **general Huyghen's theorem** then states that the residual square mean with respect to $\mathcal{H}$ can be decomposed into the sum of the residual square mean with respect to a same-dimensional parallel space $\mathcal{H}'$ through the barycenter $G$ of $M^J$ and the squared distance from $G$ to $\mathcal{H}$: $$\sum_{j\in J}f_j(M^jH^j)^2=\sum_{j\in J}f_j(M^jA^j)^2+(GG')^2,$$ where $G'$ is the projection of $G$ onto $\mathcal{H}$.

```{r, echo=TRUE}
huyghen.general.test(cloud = Target, weights = T_freq, subspace = axis)
```

### 3.2.3 Fitted and Residual Clouds

When a projected cloud $A^J$ (in a subspace $\mathcal{A}$) is used to model its preimage $M^J$, it is called a *fitted cloud*, and the vectors $\overrightarrow{A^jM^j}$, whose lengths are the residuals mentioned above, are called *residual deviations*. This leads to the geometric data--model--error formulation $$\forall j\in J\ \ M^j=A^j+\overrightarrow{A^jM^j},$$ where each $\overrightarrow{A^jM^j}$ is orthogonal to $\mathcal{A}$. Consequently, the cloud $R^J$ consisting of points $R^j=G'+\overrightarrow{H^jM^j}$ (where $G'$, as above, is the barycenter of $H^J$) depends only on the choice of $\mathcal{H}$. $R^J$ is called the *residual cloud*, and it is constructed so that $G$ is its barycenter. This yields the orthogonal decomposition $$\overrightarrow{GM^j}=\overrightarrow{GA^j}+\overrightarrow{GR^j}.$$

It is straightforward to check that the variance of $R^J$ equals the residual square mean of $M^J$ with respect to $\mathcal{A}$. The distance formula, applied pointwise to the orthogonal decomposition, then yields the variance decomposition $${Var}M^J={Var}A^J+{Var}R^J$$ into the cloud's *total variance*, its *fitted variance*, and its *residual variance*.

```{r, echo=TRUE}
cloud.decomposition(cloud = Target, subspace = axis)
cloud.variance.test(cloud = Target, weights = T_freq, subspace = axis)
```

### 3.2.4 Variables Attached to an Axis

The linear formulation in the next section relies on several coordinatization schemes. Begin with an axis $(G,\overrightarrow{\alpha})$, through the barycenter of $M^J$ in the direction of $\overrightarrow{\alpha}$, and the projected cloud $A^J$ onto this axis. We then define four centered (having mean point zero) variables:

- The *covariant variable* $\alpha^J=(\alpha^j)_{j\in J}$ has *covariant coordinates* $\alpha^j=\langle\overrightarrow{GM^j}\mid\overrightarrow{\alpha}\rangle$ and variance ${Var}A^J\|\overrightarrow{\alpha}\|^2$. The term refers to the fact that these coordinates scale with $\overrightarrow{\alpha}$.
- The *calibrated variable* $y^J=(y^j)_{j\in J}$ has *calibrated coordinates* $y^j=\alpha^j/\|\overrightarrow{\alpha}\|$ and variance ${Var}A^J$.
- The *standard variable* $z^J=(z^j)_{j\in J}$ has *standard coordinates* $z^j=y^j/SDy^J$ and variance 1.
- The *axial variable* $t^J=(t^j)_{j\in J}$ has *axial coordinates* $t^j=\langle\overrightarrow{GM^j}\mid\overrightarrow{\alpha}\rangle/\|\overrightarrow{\alpha}\|^2$ and variance ${Var}t^J=A^J/\|\overrightarrow{\alpha}\|^2$. The $t^j$ are the projections---*GDA* calls them the coordinates---of the $A^j$ onto $(G,\overrightarrow{\alpha})$.

```{r}

```